---
title: "Plotting and Color in R"
author: "Andrew Witherspoon"
date: "9/30/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

dir <- "/Users/andrew/Course4Week3"
setwd(dir)
```

The colorRamp function gives the RBG blends between two given colors - here we'll do red and and blue:
[,1] is red
[,2] is blue
[,3] is green
```{r}
pal <- colorRamp(c("red", "blue"))
```
pal(0) will be full red:
```{r}
pal(0)
```
pal(1) will be full blue:
```{r}
pal(1)
```
pal(.5) will be a blend of red and blue (purple):
```{r}
pal(.5)
```

If you give this pal() a sequence of numbers between 0 and 1, it will return a sequence of colors between red and blue:
```{r}
pal(seq(0, 1, len = 10))
```

Another option is using the colorRampPalette(), which will make your pal() function output hex color codes:
```{r}
pal <- colorRampPalette(c("red", "yellow"))
```
If you do pal(2), it will output a vector of two colors, full red and full yellow:
```{r}
pal(2)
```
If you do pal(10), it will output a vector of 10 colors - the first is full red, the last is full yellow.  Then 8 in-between colors:
```{r}
pal(10)
```

The RColorBrewer package has pre-made pallettes.  To get three colors from a RColorBrewer pallet, do this:
```{r}
library(RColorBrewer)
cols <- brewer.pal(3, "BuGn")
cols
```
Now we can use colorRampPalette() on our 3 colors (stored in cols) to create a smooth color ramp.  We'll use it on our volcano height matrix:
```{r}
pal <- colorRampPalette(cols)
image(volcano, col = pal(20))
```

The smoothScatter() function is usefull when you have a scatterplot with a lot of overlapping information, so that it is more readable.  It essentially uses a histogram to make colors darker where there are lots of points:
```{r}
par(mfrow = c(1,2))
x <- rnorm(10000)
y <- rnorm(10000)
plot(x,y)
smoothScatter(x,y)
par(mfrow = c())
```

The other option, of course, is to use transparency:
```{r}
plot(x,y, col = rgb(0,0,0,0.2), pch = 19) #0.2 is the alpha
```